Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-09-17 pot/LaTeX/skewnormal.tex
manuscript updates

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z_t$ and $\sigma^2_t$ terms.
We handle this problem with a daily random partition similar to \citet{Huser2014} that allows $z_t$ and $\sigma^2_t$ to vary by site.
The model then becomes
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + \sigma_t(\bs) v_t(\bs).
\end{align}
In this extension of (\ref{eq:fullmodel}), $z_t(\bs)$ and $\sigma_t(\bs)$ are allowed to vary by site.
Their spatial variation is determined by the random partition model defined below.
Consider a set of daily spatial knots $\bw_{tk} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{t1}, \ldots, P_{tK}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{tk} = \{ \bs : k = \argmin_\ell || \bs - \bw_{t\ell} || \}.
\end{align*}
So, for $\bs \in P_{tk}$, let
\begin{align}
  z_t(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2_t(\bs) &= \sigma^2_{tk} \label{eq:sitesig}.
\end{align}
Then within each partition, $Y_t(\bs)$ follows the distribution given in (\ref{eq:fullmodel}).
When incorporating the random daily partition, the $chi$ statistic becomes
\begin{align}
  \chi(\bh) = \lim_{c \rightarrow \infty} \pi(\bh) \chi(\bh)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(\bh) = \lim_{h \rightarrow \infty} \pi(\bh) \chi(\bh) = 0.
\end{align}
A proof of this can be found in Appendix A.3.

@@

\subsection{Design}\label{s:simdesign}
For all simulation designs, we generate data from the model presented in Section \ref{s:part} using $n_s=130$ sites and $n_t=50$ independent days.
The sites are generated Uniform$([0, 10] \times [0, 10])$.
We generate data from 6 different simulation designs:
\begin{enumerate} \setlength{\itemsep}{-0.5em}
  \item Gaussian marginal, $K=1$ knot
  \item $t$ marginal, $K=1$ knot
  \item $t$ marginal, $K=5$ knots
  \item skew-$t$ marginal, $K=1$ knots
  \item skew-$t$ marginal, $K=5$ knots
  \item Max-stable.
\end{enumerate}
In the first five designs, the $v_t(\bs)$ terms are generated using a \Matern covariance with smoothness parameter, $\nu = 0.5$, and spatial range, $\rho = 0.1$.
For the covariance matrices in designs 1 -- 5, the proportion of the variance accounted for by the spatial variation is $\gamma = 0.9$ while the proportion of the variance accounted for by the nugget effect is $0.1$.
In the first design, $\sigma^2 = 2$ is used for all days.
For designs 2 -- 4, $sigma^2_{tk} \iid \text{IG}(3, 8)$
For designs 1 -- 3, we set $\alpha = 0$.
For designs four and five, $\alpha = 3$ was used, and the $z_t$ are generated as described in (\ref{eq:sitez}).
In the sixth design, we generate from a spatial max-stable distribution \citep{Reich2012} with parameters $\mu = 1, \sigma=1, \xi=0.2$ and 144 spatial knots on a regular lattice in the square $[1, 9] \times [1, 9]$.
In all six designs, the mean ($\mu(\bs)$) is assumed to be constant across space.

@@

\section{Data analysis}\label{s:analysis}
To illustrate this method, we consider the daily maximum 8-hour ozone measurements for July 2005 at 735 Air Quality System (AQS) monitoring sites in the eastern United States as the response.
For each site, we also have covariate information containing the estimated ozone from the Community Multi-scale Air Quality (CMAQ) modeling system.
Initially, we fit a linear regression assuming that the mean function can be expressed as
\begin{align}
  \mu_t(\bs) = \beta_0 + \beta_1 \cdot \text{CMAQ}_t(\bs). \label{eq:datamean}
\end{align}
The data from 5 July are shown in Figure \ref{fig:ozone} along with a Q-Q plot of the residuals compared to a skew-$t$ distribution with 8 d.f. and $\alpha = 1$.
\begin{center}
\begin{figure}
  \includegraphics[width=0.5\linewidth]{plots/ozone-5jul.pdf}
  \includegraphics[width=0.5\linewidth]{plots/qq-res.pdf}
  \caption{Ozone values on 5 July 2005 (left) Q-Q plot of the residuals (right)}
  \label{fig:ozone}
\end{figure}
\end{center}
The $\chi(h)$ and $\chi(t)$ plots indicate that the residuals still demonstrate spatial and temporal dependence (see Figure \ref{fig:chi-st})
\begin{center}
\begin{figure}
  \includegraphics[width=0.5\linewidth]{plots/chi-h-ozone.pdf}
  \includegraphics[width=0.5\linewidth]{plots/chi-t-ozone.pdf}
  \caption{$\chi(h)$ plot for the residuals. The vertical line indicates distance after which observations no longer demonstrate dependence (left) $\chi(t)$ plot for the residuals (right)}
  \label{fig:chi-st}
\end{figure}
\end{center}

@@

\section*{Appendix A.1: MCMC details}
The MCMC sampling for the model \ref{s:hier} is done using {\tt R} (http://www.r-project.org). Whenever possible, we select conjugate priors (see Appendix A.2); however, for some of the parameters, no conjugate prior distributions exist.
When no conjugate prior distribution exists, we use a random walk Metropolis Hastings update step.
In each Metropolis Hastings update, we tune the algorithm to give acceptance rates near 0.40.d

\subsection*{Spatial knot locations}
For each day, we update the spatial knot locations, $\bw_1, \ldots, \bw_K$, using a Metropolis Hastings block update.
First, we generate candidate knots locations using a random walk Gaussian candidate distribution
\begin{align*}

\end{align*}

\subsection*{Spatial covariance parameters}
We update the three spatial covariance parameters, $\log(\rho)$, $\log(\nu)$, $\gamma$, using a Metropolis Hastings block update step.
First, we generate a candidate using a random walk Gaussian candidate distribution 
\begin{align*}
	\log(\rho)^{(c)} \sim \text{N}(\log(\rho)^{(r - 1)}, s^2)
\end{align*}
where $\log(\rho)^{(r-1)}$ is the value at MCMC iteration $r - 1$, and $s$ is a tuning parameter.
Candidates are generated for $\log(\nu)$ and $\gamma$ in a similar fashion.
The acceptance ratio is 
\begin{align*}
	R = \left\{ \frac{ \prod_{t = 1}^{T} l[Y_t(\bs | \rho^{(c)}, \nu^{(c)}, \gamma^{(c)}, \ldots] }{\prod_{t = 1}^{T} l[Y_t(\bs | \rho^{(r-1)}, \nu^{(r-1)}, \gamma^{(r-1)}, \ldots] } \right\} \times \left\{ \frac{ p[\rho^{(c)}] }{ p[\rho^{(r - 1)] } } \right\} \times \left\{ \frac{ p[\nu^{(c)}] }{ p[\nu^{(r - 1)}] } \right\} \times \left\{ \frac{ p[ \gamma^{(c)} ] }{ p[\gamma^{(r - 1)} ] } \right\}
\end{align*}
where $l$ is the likelihood given in \ref{eq:hier}.
All three candidates are accepted with probability $\min\{R, 1\}$.

>>>>

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z_t$ and $\sigma^2_t$ terms.
We handle this problem with a daily random partition similar to \citet{Huser2014} that allows $z_t$ and $\sigma^2_t$ to vary by site.
The model then becomes
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + \sigma_t(\bs) v_t(\bs).
\end{align}
In this extension of (\ref{eq:fullmodel}), $z_t(\bs)$ and $\sigma_t(\bs)$ are allowed to vary by site.
Their spatial variation is determined by the random partition model defined below.
Consider a set of daily spatial knots $\bw_{tk} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{t1}, \ldots, P_{tK}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{tk} = \{ \bs : k = \argmin_\ell || \bs - \bw_{t\ell} || \}.
\end{align*}
So, for $\bs \in P_{tk}$, let
\begin{align}
  z_t(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2_t(\bs) &= \sigma^2_{tk} \label{eq:sitesig}.
\end{align}
Then within each partition, $Y_t(\bs)$ follows the distribution given in (\ref{eq:fullmodel}).
When incorporating the random daily partition, the $\chi$ statistic becomes
\begin{align}
  \chi(\bh) = \lim_{c \rightarrow \infty} \pi(\bh) \chi(\bh)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(\bh) = \lim_{h \rightarrow \infty} \pi(\bh) \chi(\bh) = 0.
\end{align}
A proof of this can be found in Appendix A.3.

@@

\subsection{Design}\label{s:simdesign}
For all simulation designs, we generate data from the model presented in Section \ref{s:part} using $n_s=130$ sites and $n_t=50$ independent days.
The sites are generated Uniform$([0, 10] \times [0, 10])$.
We generate data from 6 different simulation designs:
\begin{enumerate} \setlength{\itemsep}{-0.5em}
  \item Gaussian marginal, $K=1$ knot
  \item $t$ marginal, $K=1$ knot
  \item $t$ marginal, $K=5$ knots
  \item skew-$t$ marginal, $K=1$ knots
  \item skew-$t$ marginal, $K=5$ knots
  \item Max-stable.
\end{enumerate}
In the first five designs, the $v_t(\bs)$ terms are generated using a \Matern covariance with smoothness parameter, $\nu = 0.5$, and spatial range, $\rho = 0.1$.
For the covariance matrices in designs 1 -- 5, the proportion of the variance accounted for by the spatial variation is $\gamma = 0.9$ while the proportion of the variance accounted for by the nugget effect is $0.1$.
In the first design, $\sigma^2 = 2$ is used for all days.
For designs 2 -- 4, $\sigma^2_{tk} \iid \text{IG}(3, 8)$
For designs 1 -- 3, we set $\alpha = 0$.
For designs four and five, $\alpha = 3$ was used, and the $z_t$ are generated as described in (\ref{eq:sitez}).
In the sixth design, we generate from a spatial max-stable distribution \citep{Reich2012} with parameters $\mu = 1, \sigma=1, \xi=0.2$ and 144 spatial knots on a regular lattice in the square $[1, 9] \times [1, 9]$.
In all six designs, the mean ($\mu(\bs)$) is assumed to be constant across space.

@@

\section{Data analysis}\label{s:analysis}
To illustrate this method, we consider the daily maximum 8-hour ozone measurements for July 2005 at 735 Air Quality System (AQS) monitoring sites in the eastern United States as the response.
For each site, we also have covariate information containing the estimated ozone from the Community Multi-scale Air Quality (CMAQ) modeling system.
Initially, we fit a linear regression assuming that the mean function can be expressed as
\begin{align}
  \mu_t(\bs) = \beta_0 + \beta_1 \cdot \text{CMAQ}_t(\bs). \label{eq:datamean}
\end{align}
The data from 10 July are shown in Figure \ref{fig:ozone} along with a Q-Q plot of the residuals compared to a skew-$t$ distribution with 8 d.f. and $\alpha = 1$.
\begin{center}
\begin{figure}
  \includegraphics[width=0.5\linewidth]{plots/ozone-10jul.pdf}
  \includegraphics[width=0.5\linewidth]{plots/qq-res.pdf}
  \caption{Ozone values on 10 July 2005 (left) Q-Q plot of the residuals (right)}
  \label{fig:ozone}
\end{figure}
\end{center}
The $\chi(h)$ and $\chi(t)$ plots indicate that the residuals still demonstrate spatial and temporal dependence (see Figure \ref{fig:chi-st})
\begin{center}
\begin{figure}
  \includegraphics[width=0.5\linewidth]{plots/chi-h-ozone.pdf}
  \includegraphics[width=0.5\linewidth]{plots/chi-t-ozone.pdf}
  \caption{$\chi(h)$ plot for the residuals. The vertical line indicates distance after which observations no longer demonstrate dependence (left) $\chi(t)$ plot for the residuals (right)}
  \label{fig:chi-st}
\end{figure}
\end{center}

@@

\section*{Appendix A.1: MCMC details}
The MCMC sampling for the model \ref{s:hier} is done using {\tt R} (http://www.r-project.org). Whenever possible, we select conjugate priors (see Appendix A.2); however, for some of the parameters, no conjugate prior distributions exist.
When no conjugate prior distribution exists, we use a random walk Metropolis Hastings update step.
In each Metropolis Hastings update, we tune the algorithm to give acceptance rates near 0.40.

\subsection*{Spatial knot locations}
For each day, we update the spatial knot locations, $\bw_1, \ldots, \bw_K$, using a Metropolis Hastings block update.
Let $\Phi_2$ represent a bivariate normal density function, and let $\bw^*_k = h[\Phi_2^{-1}(\bs_k)]$ where $h(\cdot)$ is a transformation to shift and scale the transformed knots so that $\bs^*_k \in \calD$.
Because the spatial domain is bounded, we generate candidate knots using $\bw^*_1, \ldots, \bw^*_K$ and a random walk bivariate Gaussian candidate distribution
\begin{align*}
	{\bw^*_k}^{(c)} \sim \text{N}({\bw^*_k}^{(r - 1)}, s^2 I_2)
\end{align*}
where $\bw*_k^{(r - 1)}$ is the location for the transformed knot at MCMC iteration $r - 1$, $s$ is a tuning parameter, and $I_2$ is an identity matrix.
After candidates have been generated for all $K$ knots, the acceptance ratio is
\begin{align*}
  R = \left\{ \frac{ l[ Y_t(\bs | \bw_1^{(c)}, \ldots, \bw_K^{(c)}, \ldots)] }{l[ Y_t(\bs | \bw_1^{(r - 1)}, \ldots, \bw_K^{(r - 1)}, \ldots)]} \right\} \times \left\{ \frac{ \prod_{k = 1}^{K}\phi(\bw_k^{(c)})}{ \prod_{k = 1}^{K}\phi(\bw_k^{(r - 1)})} \right\}
\end{align*}
where $l$ is the likelihood given in (\ref{eq:hier}).
The candidate knots are accepted with probability $\min\{R, 1\}$.

\subsection*{Spatial covariance parameters}
We update the three spatial covariance parameters, $\log(\rho)$, $\log(\nu)$, $\gamma$, using a Metropolis Hastings block update step.
First, we generate a candidate using a random walk Gaussian candidate distribution
\begin{align*}
	\log(\rho)^{(c)} \sim \text{N}(\log(\rho)^{(r - 1)}, s^2)
\end{align*}
where $\log(\rho)^{(r-1)}$ is the value at MCMC iteration $r - 1$, and $s$ is a tuning parameter.
Candidates are generated for $\log(\nu)$ and $\gamma$ in a similar fashion.
The acceptance ratio is
\begin{align*}
	R = \left\{ \frac{ \prod_{t = 1}^{T} l[Y_t(\bs | \rho^{(c)}, \nu^{(c)}, \gamma^{(c)}, \ldots] }{\prod_{t = 1}^{T} l[Y_t(\bs | \rho^{(r-1)}, \nu^{(r-1)}, \gamma^{(r-1)}, \ldots] } \right\} \times \left\{ \frac{ p[\rho^{(c)}] }{ p[\rho^{(r - 1)] } } \right\} \times \left\{ \frac{ p[\nu^{(c)}] }{ p[\nu^{(r - 1)}] } \right\} \times \left\{ \frac{ p[ \gamma^{(c)} ] }{ p[\gamma^{(r - 1)} ] } \right\}.
\end{align*}
All three candidates are accepted with probability $\min\{R, 1\}$.
